Proportional multi-state multiple-cohort life table model

### 2018-04-09 MSLT/MSLTdoc.Rmd
GBD data inputs

#The work in this document requires the following packages (code, rmarkdown and citing packages):
require(dplyr)
require(tidyverse)
require(knitr)
require(kableExtra)
require(citr)
```
# Introduction

@@

![Basic ITHIMR infrastructure](structure\Figure1.png){width=70%, fig.pos="H"}

@@

@@

* **Interaction between multiple diseases**
    The *PMSLT* accounts for the interaction between multiple diseases, with proportions of the population        being able to be in more than one health state [@RN8957]. This avoids  overestimation of outcomes as a        result of summing health outcomes attributable to each disease individually as done in *ITHIM*. It is         important to note that the *PMSLT* assumes that diseases are independent of each other. That is to            say, developing a disease is unrelated to a concurrent diagnoses of another disease).

* **Mortality rate** 
    The *PMSLT* calculations for changes in life years (and health-adjusted life years) and mortality outcomes     is based on observed mortality rates for the population of interest. In the *ITHIM* model, if burden of       disease estimates from the Global Burden of Disease (GBD) study are used, then, the mortality component is     based on the highest attained life expectancy observed in the world.

* **Impact of disability in increased life expectancy**
    In GBD studies, YLLs are not adjusted for disability; hence, their use in estimating intervention effects     results in over-estimation, which the *PMSLT* approach avoids. Another way of seeing this is that             estimated changes in morbidity using the *ITHIM* do not allow for how implicit increases in life              expectancy impact on morbidity. While the changes in deaths and prevalence using the *PMSLT* are in some      ways more accurate than those from the *ITHIM* approach it should be noted that that the average age of       death and incident disease will change and thus the disease burden will be on average be shifted later in     life (which is a realistic approach).

@@

(NEED TO PREPARE: LONDON, SAO PABLO, DELHI, BAY AREA). DATA AVAILABILITY FROM ARE IN 5 YEARS UP TO 95+. gbd DATA IS NOT AVAILABLE AT THE CITY LEVEL, HOWEVER, SOME LEVEL OF DISSAGREAGATION IS AVAILABLE. ENGLAND: EAST MIDLANDS, EAST OF ENGLAND, GRETER LONDON, ETC. wE COULD USE GREATER LONDON. FOR EDINBURGH WE HAVE DATA FOR SCOTLAND.  FOR SAO PABLO, DATA IS FOR BRAZIL. MEXICO IS NOT LISTED, BUT GBD HAS DATA FOR DISTRITO FEDERAL. FOR DELHI THE DATA IS FOR INDIA. FOR ACCRA THE DATA IS FOR GHANA. FOR LOS ANGELES DATA IS FOR UNITED STATES, BUT WE CAN HAVE THIS FROM NEIL?

@@

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**. The color coding from Data inputs sources to Inputs PMSLT link sources with inputs for the PMSLT. Solid arrows represent final inputs and dashed-arrows represent intermediate inputs that need further processing. Purple coding means a process and green coding represent change in mortality and disability prevalence rates to modify the life table parameters. Black color coding with white color font represent final model outcomes. For both, the life table and disease life table, two sets of each are simulated, one for the baseline and the other for the scenario. **Data inputs** are from four sources: Sythetic population, GBD data (from the Institute of Health Metrics and Evaluations (IHME)), ITHIM calculations and Injuries model. The synthetic population provides population numbers by age-groups and sex which are used in the life table (**PMSLT Mechanisms**) with each age cohort simulated until they die or reach 100 years of age. Population numbers are also used to derive road death rates and road injuries rates from numbers for the baseline and scenario generated by the Injuries model. The difference in road deaths rates modifies all cause-mortatliy in the scenario life table and the change in injuries rates modifies total YLDs rates (hence, healthy life years). Data from GBD IHME includes: total YLD rates per one, disease specific YLDs (numbers) and prevalence to derive disability weights for the included diseases, all cause mortatliy rates per one, population numbers and diseases parameters (incidence, mortality and prevalence). Dismod II is used to derive internally consitent estimates incidence and case fatality rates from GBD IHME data, which are inputs of the disease specific life tables. Case fatality derive from Dismod II is the same for the baseline and scenario (however, could be changed) and incidence is different. For the baseline life tables incidence is as derived from Dismod II and for the scanrio is modified by the PIF (1-PIF). A change in incidence modifies prevalence and mortality in later years. The difference between prevalence and mortality rates between the baseline and scanrio then feed into the scenario life table.* ](structure\Figure3.png){width=70%, hight=50%, fig.pos="H"}

@@

>>>>

#The work in this document requires the following packages (code, rmarkdown and citing packages):
require(dplyr)
require(tidyverse)
require(knitr)
require(kableExtra)
require(citr)
```

\newpage

# Introduction

@@

@@

\newpage

![Basic ITHIMR infrastructure](structure\Figure1.png){fig.pos="H"}

@@

***Interaction between multiple diseases**
The *PMSLT* accounts for the interaction between multiple diseases, with proportions of the population    being able to be in more than one health state [@RN8957]. This avoids  overestimation of outcomes as a    result of summing health outcomes attributable to each disease individually as done in *ITHIM*. It is     important to note that the *PMSLT* assumes that diseases are independent of each other. That is to        say, developing a disease is unrelated to a concurrent diagnoses of another disease).

***Mortality rate** 
The *PMSLT* calculations for changes in life years (and health-adjusted life years) and mortality outcomes is based on observed mortality rates for the population of interest. In the *ITHIM* model, if burden of  disease estimates from the Global Burden of Disease (GBD) study are used, then, the mortality component is based on the highest attained life expectancy observed in the world.

***Impact of disability in increased life expectancy**
In GBD studies, YLLs are not adjusted for disability; hence, their use in estimating intervention effects results in over-estimation, which the *PMSLT* approach avoids. Another way of seeing this is that         estimated changes in morbidity using the *ITHIM* do not allow for how implicit increases in life          expectancy impact on morbidity. While the changes in deaths and prevalence using the *PMSLT* are in some  ways more accurate than those from the *ITHIM* approach it should be noted that that the average age of   death and incident disease will change and thus the disease burden will be on average be shifted later in life (which is a realistic approach).

@@

Initial case studies for the ITHIMR are: London, Sao PAblo, Delhi, Accra, Los Angeles and Edinburgh. Here, we will start with **greater London** given the availability of dieases epidemiology data from the GBD IHME study. For the rest of the case study cities data is available at the country level, hence, a scaling method is needed.

@@

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**. The color coding from Data inputs sources to Inputs PMSLT link sources with inputs for the PMSLT. Solid arrows represent final inputs and dashed-arrows represent intermediate inputs that need further processing. Purple coding means a process and green coding represent change in mortality and disability prevalence rates to modify the life table parameters. Black color coding with white color font represent final model outcomes. For both, the life table and disease life table, two sets of each are simulated, one for the baseline and the other for the scenario. **Data inputs** are from four sources: Sythetic population, GBD data (from the Institute of Health Metrics and Evaluations (IHME)), ITHIM calculations and Injuries model. The synthetic population provides population numbers by age-groups and sex which are used in the life table (**PMSLT Mechanisms**) with each age cohort simulated until they die or reach 100 years of age. Population numbers are also used to derive road death rates and road injuries rates from numbers for the baseline and scenario generated by the Injuries model. The difference in road deaths rates modifies all cause-mortatliy in the scenario life table and the change in injuries rates modifies total YLDs rates (hence, healthy life years). Data from GBD IHME includes: total YLD rates per one, disease specific YLDs (numbers) and prevalence to derive disability weights for the included diseases, all cause mortatliy rates per one, population numbers and diseases parameters (incidence, mortality and prevalence). Dismod II is used to derive internally consitent estimates incidence and case fatality rates from GBD IHME data, which are inputs of the disease specific life tables. Case fatality derive from Dismod II is the same for the baseline and scenario (however, could be changed) and incidence is different. For the baseline life tables incidence is as derived from Dismod II and for the scanrio is modified by the PIF (1-PIF). A change in incidence modifies prevalence and mortality in later years. The difference between prevalence and mortality rates between the baseline and scanrio then feed into the scenario life table.*](structure\Figure3.png){width=50%,hight=30%,fig.pos="H"}

@@

####Global Burden of Disease data

First, we explain how to obtain the data, second additional processing to derive data not reported (population) and one-year age groups (original data is in five-year age groups) and last procedure to use Dismod II. Data from the *Global Burden of Disease data* in Figure 3 can be download from here: http://ghdx.healthdata.org/gbd-results-tool. Figure 4 is a screenshot of the GHDx.

![Global Burden of Disease data results tool.](data/Figure4.png){width=70%,hight=50%,fig.pos="H"}

Table 2 specifies the selections to do for each of the tabs in Figure 2.

```{r, echo=FALSE}
pmslt_inputs <- tribble(
~Tab, ~Selection,
"Base", "Single", "Location", "Case study city", "Year", "Latest available", "Context", "Cause", "Age", "Under 5, 5 to 9, 10 to 14, 15 to 19, 20 to 24, 25 to 29, 30 to 34, 35 to 39, 40 to 49, 50 to 54, 55 to 59, 60 to 64, 65 to 69, 70 to 74, 75 to 79, 80 to 84, 89 to 89, 90 to 95, 95 plus", "Metric", "Number, Rate", "Measure", "Deaths, YLDs, Prevalence, Incidence", "Sex", "Male, Female", "Cause", "Total All causes, ischemic heart disease, etc")

kable(pmslt_inputs,"latex", booktabs = T, caption = "Global burden of disease data") %>%
 kable_styling(latex_options = c("striped", "hold position"),
               full_width = F)%>%
  column_spec(2, width = "15em") %>%
  column_spec(3, width = "15em") %>%
  row_spec(0, bold = T)
```

Health outcomes included in leisure time meta-analysis: all-cause mortality, cardiovascular disease, stroke, coronary heart disease, total cancer, colon, lung, endometrial, and breast cancer.
